---
title: "Methylation_box+bar"
output: html_document
date: "2026-03-22"
---

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath("/home1/spoden/490_cluster/analysis_data/"))
```

```{r}
#if (!require("DESeq2", quietly = TRUE))
  #BiocManager::install("DESeq2")

library(DESeq2)
library(ggplot2)
```

```{r}
# Load all data
rna_clinical <- read.csv("brca_rna_clinical_data.csv", row.names = 1)
rna_genes    <- read.csv("brca_rna_gene_data.csv")
rna_counts   <- read.csv("brca_rna_count_data.csv", row.names = 1)

methylation_clinical <- read.csv('brca_methylation_clinical.csv', row.names = 1)
betas <- read.csv('brca_methylation_betas.csv', row.names = 1)
cpg_sites <- read.csv('brca_cpg_sites.csv', row.names = 1)
dat <- read.csv('methylation_results_metastasis.csv', row.names = 1)
```

```{r}
# Add metastatic_status to methylation_clinical
methylation_clinical$metastatic_status <- ifelse(
  methylation_clinical$definition == "Metastatic", "Metastatic", "Non-Metastatic"
)

# Filter methylation data to Metastatic and Primary solid Tumor only
met_mask <- methylation_clinical$definition %in% c("Metastatic", "Primary solid Tumor")
methylation_clinical <- methylation_clinical[met_mask, ]
betas_clean <- betas[, met_mask]
```

```{r}
# DESeq2 preprocessing
definition_mask <- rna_clinical$definition %in% c("Metastatic", "Primary solid Tumor")
rna_clinical <- rna_clinical[definition_mask, ]
rna_counts   <- rna_counts[, definition_mask]

rna_clinical$metastatic_status <- factor(
  ifelse(rna_clinical$definition == "Metastatic", "Metastatic", "Non-Metastatic"),
  levels = c("Non-Metastatic", "Metastatic")
)

rna_clinical$age_category <- factor(
  ifelse(rna_clinical$age_at_index >= 58, "old", "young")
)

na_mask      <- !is.na(rna_clinical$age_at_index)
rna_clinical <- rna_clinical[na_mask, ]
rna_counts   <- rna_counts[, na_mask]

keep_genes         <- rowSums(rna_counts) >= 20
rna_counts         <- rna_counts[keep_genes, ]
rna_genes_filtered <- rna_genes[keep_genes, ]

colnames(rna_counts) <- gsub("\\.", "-", colnames(rna_counts))
```

```{r}
# Run DESeq2
dds <- DESeqDataSetFromMatrix(countData = rna_counts,
                              colData   = rna_clinical,
                              design    = ~ age_category + metastatic_status)
dds_obj <- DESeq(dds)

results <- results(dds_obj,
                   format   = "DataFrame",
                   contrast = c("metastatic_status", "Metastatic", "Non-Metastatic"))

results <- data.frame(results)
results$gene_name <- rna_genes_filtered$gene_name

write.csv(results, "deSeq_results_metastatic.csv", row.names = TRUE)
deseq_results <- read.csv("deSeq_results_metastatic.csv")
```

```{r}
# Find genes of interest - downregulated AND hypermethylated in metastatic
downregulated <- deseq_results[deseq_results$padj < 0.05 &
                                deseq_results$log2FoldChange < -1, 'gene_name']
hypermethylated <- dat[dat$foldchange > 1, 'geneName']
interest_genes <- intersect(downregulated, hypermethylated)
print(interest_genes)  
```

```{r}
GENE <- "MMP14"

gene_counts_mask <- rna_genes$gene_name == GENE
gene_betas_mask  <- cpg_sites$gene == GENE

# Group masks
rna_meta_mask    <- rna_clinical$metastatic_status == "Metastatic"
rna_nonmeta_mask <- rna_clinical$metastatic_status == "Non-Metastatic"

meth_meta_mask    <- methylation_clinical$metastatic_status == "Metastatic"
meth_nonmeta_mask <- methylation_clinical$metastatic_status == "Non-Metastatic"

# Extract data
rna_meta    <- as.numeric(rna_counts[gene_counts_mask, rna_meta_mask])
rna_nonmeta <- as.numeric(rna_counts[gene_counts_mask, rna_nonmeta_mask])

methylation_meta    <- betas_clean[gene_betas_mask, meth_meta_mask]
methylation_nonmeta <- betas_clean[gene_betas_mask, meth_nonmeta_mask]
```

```{r}
# Boxplot - RNA expression
boxplot(rna_nonmeta, rna_meta,
        xlab = 'Group', ylab = 'Counts',
        names = c('Non-Metastatic', 'Metastatic'),
        col = c("blue", "red"),
        main = paste("RNA Expression of", GENE, ": Metastatic vs Non-Metastatic BRCA"))
```

```{r}
# Barplot - CpG site methylation
plot(1:dim(methylation_nonmeta)[1], apply(methylation_nonmeta, 1, mean),
     col = 'blue', type = 'h', lwd = 3,
     ylab = 'Average Beta', xlab = 'CpG sites',
     main = paste("CpG Methylation at", GENE, ": Metastatic vs Non-Metastatic BRCA"))

lines(1.3:(dim(methylation_meta)[1]+0.3), apply(methylation_meta, 1, mean),
      col = 'red', type = 'h', lwd = 3)

legend("topright",
       legend = c("Non-Metastatic", "Metastatic"),
       col = c("blue", "red"),
       lwd = 3,
       cex = 0.7)
```

Analysis:
The boxplot shows that MMP14 RNA expression is greater in metastatic patients than non-metastatic patients. This is unexpected because the gene was in the hypermethylated and downregulated gene list. The CpG methylation barplot shows mixed results as some sites show higher methylation in metastatic patients and some show it in non-metastatic patients. This suggests that there is no consistent hypermethylation pattern across the gene. Since these results are contradictory, we can't conclude that MMP14 transcription is being regulated by DNA methylation. 
